Variance reduction for MCMC methods via martingale representations

### 2020-01-28 aap/Revision/vr-mcmc-revision.tex
fixed misprint

We shall use the notations $\mathbb N=\{1,2,\ldots\}$ and $\mathbb N_0=\mathbb N\cup\{0\}$.
By $H_k$, $k\in\mathbb N_0$,
we denote the normalized Hermite polynomial on $\mathbb R$, that is,
$$
H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
\quad x\in\mathbb R.
$$
For a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
$\mathbf{H}_\mathbf{k}$ denotes the normalized Hermite polynomial on $\mathbb R^d$, that is,
$$
\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
$$
In what follows, we also use the notation
$|\mathbf{k}|=\sum_{i=1}^d k_i$ for $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
and we set $\mathcal G_p=\sigma(Z_1,\ldots,Z_p)$, $p\in\mathbb N$, and $\mathcal G_0=\mathrm{triv}$.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{align}
M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{p=N+1}^{N+n}\omega_{p,n}^{N}\sum_{l=N+1}^{p}a_{p,l,\mathbf{k}}(X_{l-1})\mathbf{H}_\mathbf{k}(Z_{l})
\label{eq:29032018a5}\\
&=\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{l=N+1}^{N+n}\bar{a}_{l,\mathbf{k}}(X_{l-1})\mathbf{H}_\mathbf{k}(Z_{l})
\notag
\end{align}
with \(\|\mathbf{k}\|=\max_{i} k_i\) and
\begin{align*}
\bar{a}_{l,\mathbf{k}}(x)
& =\sum_{p=l}^{N+n}\omega_{p,n}^{N}a_{p,l,\mathbf{k}}(x) =\mathsf{E}\left[\left.\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right)\mathbf{H}_\mathbf{k}\left(Z_{l}\right)\right|X_{l-1}=x\right].
\end{align*}
For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we shall be interested in the Mean Squared Error (MSE),
which can be decomposed as the sum of the squared  bias
and the  variance:
\begin{eqnarray}\label{eq:29032018a4}
\mathrm{MSE}\left[\rho(f)\right]
&=&\mathsf{E}\left[(\rho(f)-\pi(f))^2\right]
\\
\nonumber
&=&\left(\mathsf{E}[\rho(f)]-\pi(f)\right)^2
+\mathsf{Var}[\rho(f)].
\end{eqnarray}
Our analysis will be done under the following two assumptions.

\begin{description}
\item[(H1)][\textbf{Lipschitz continuity}] Assume that $\nabla U$ is Lipschitz, that is, there exists $L_U\geq 0$ such that 
|\nabla U(x)-\nabla U(y)|\leq L_U|x-y|, \quad x,y\in \mathbb{R}^d.
\item[(H2)][\textbf{Convexity outside a ball}] There exist $K_U>0,$ $M_U>0$ and $m_U>0$ such that for any $x\not\in B_{K_U}(0)	$ and any $y\in \mathbb{R}^d$ it holds 
\langle D^2U(x)y,y\rangle \geq m_U\|y\|^2, \quad \|D^3 U(y)\| \leq M_U. 
\end{description}
First we analyse the biases of the estimators
$\pi_n^N(f)$ and $\pi_{K,n}^N(f).$
\subsection{Squared bias}
Due to the martingale transform structure
of $M_{K,n}^N(f)$,
\mathsf E\left[M_{K,n}^N(f)\right]=0,
Hence both estimators
$\pi_n^N(f)$ and $\pi_{K,n}^N(f)$ have the same conditional bias.
Notice that this remains true also when we substitute
the coefficients $a_{p,l,\mathbf{k}}$ in~\eqref{eq:29032018a5}
with some independent approximations $\widehat a_{p,l,\mathbf{k}}.$ For a bounded Borel function~$f$,
we can estimate the conditional bias
similarly to the beginning of ~\cite[Section~4]{durmus:moulines:2017}:

@@

\left(\mathsf{E}[\pi_{K,n}^N(f)]-\pi(f)\right)^2
&=\left(\mathsf{E}[\pi_{n}^N(f)]-\pi(f)\right)^2
\label{eq:06042018a2}\\
&\le
\mathrm{osc}(f)^2
\sum_{p=N+1}^{N+n}
\omega_{p,n}^N\,
\|Q_\gamma^{p}(x_0,\cdot)-\pi(\cdot)\|_{TV}^2,

@@

$\mathrm{osc}(f):=\sup_{x\in\mathbb R^d}f(x)-\inf_{x\in\mathbb R^d}f(x)$,
$\|\mu-\nu\|_{\TV}$ denotes the total variation distance
between probability measures $\mu$ and $\nu$, that is,
\|\mu-\nu\|_{TV}=\sup_{A\in\mathcal B(\mathbb R^d)}
|\mu(A)-\nu(A)|,
$\pi(\cdot)$ denotes the probability measure on $\mathbb R^d$
with density $\pi$ of~\eqref{eq:stationary_distr};
for $\gamma>0$, the Markov kernel $R_\gamma$
is defined by
R_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\}dy,
while, for $k,l\in\mathbb N$, $k\le l$,
the kernel $Q_\gamma^{k,l}$ is
$
Q_\gamma^{k,l}=R_{\gamma_l}\cdots R_{\gamma_k},
$
which, finally, provides the  measure
$Q_\gamma^{p}(x_0,\cdot)=Q_\gamma^{p,1}$
used in the right-hand side of~\eqref{eq:06042018a2}.
Different specific upper bounds for the squared bias
can be deduced from~\eqref{eq:06042018a2}
using results of Section~3 in~\cite{durmus:moulines:2017}
on bounds in the total variation distance. For example, the following proposition holds.
Suppose that the assumption (H1) holds, then it holds
\|Q_\gamma^{p}(x,\cdot)-\pi(\cdot)\|^2_{TV}\lesssim \sum_{k=r+1}^{p} \gamma_k^2+\exp\left(-\varkappa\sum_{k=r+1}^{p} \gamma_k\right),
for some $\varkappa>0,$ where $\lesssim$ stands for  inequality up to a constant depending on $L$ and $\mathcal G_N.$ In particular if $\gamma_k=k^{-\alpha}$ for some $\alpha\in (0.5,1], $ then 
\|Q_\gamma^{N+1,N+n}(x,\cdot)-\pi(\cdot)\|^2_{TV}\lesssim \frac{1}{(N+1)^{2\alpha-1}}
provided that $N/n\to 0$ as $N,n\to \infty.$
\subsection{Conditional variance}
An upper bound for the variance of the classical estimator~\eqref{eq:29032018a2}
is provided in  in~\cite[Theorem~17]{durmus:moulines:2017}.
As for the estimator~\eqref{eq:29032018a3}, 
it follows from~\eqref{eq:29032018a5}, Theorem~\ref{prop:29032018a1} and independence condition for the sequence $(Z_p)_{p \geq 1}$.
\label{cond_var_dec}
	For variance-reduced estimator~\eqref{eq:29032018a3}, defined for the Markov Chain~\eqref{eq:chain_gen} the following representation holds
\begin{equation}\label{eq:29032018a6}
\mathsf{Var}\left[\pi_{K,n}^{N}(f)\right]
%=\sum_{\min_i (k_i)\ge K+1}
=\sum_{\mathbf k\colon \|\mathbf k\|\ge K+1}
\sum_{l=N+1}^{N+n}
\mathsf{E}\left[\left|\bar a_{l,\mathbf{k}}(X_{l-1})\right|^{2}\right].
Now we present an upper bound for the right-hand sides of~\eqref{eq:29032018a6} in case of ULA algorithm.
\begin{thm}\label{th:mr}
Assume (H1) and (H2).
Suppose that $f$  is two times continuously differentiable
and it holds
|\partial^{\mathbf{k}} f(x)|\le  B_f, \quad
x\in\mathbb R^d
for all \(\mathbf{k}\) with \(0<\|\mathbf{k}\|\leq  2.\) Let $(\gamma_k)_{k\in\mathbb N}$ be a sequence of positive
and non-increasing step sizes with $\sum_{k=1}^\infty \gamma_k=\infty$ and $\sum_{k=1}^\infty \gamma^2_k<\infty.$ 
Then it holds under 
\label{eq:var-bound}
\mathsf{Var}\left[\pi_{K,n}^{N}(f)\right]\lesssim \frac{1}{\Gamma^2_{N+2,N+n+1}},

@@

where \(\lesssim\) stands for inequality up to a constant not depending  on \(n\) and \(N.\)

@@

Let us formulate and prove prove analogue of the previous theorem for inhomogeneous Markov chain with step size $\gamma_1,\ldots,\gamma_n,\ldots$. Note that under assumptions $\sum_{p=1}^{\infty} \gamma_p=\infty, \quad \sum_{p=1}^{\infty} \gamma^2_p<\infty$ and {\bf (H1)} the stationary distribution of ULA-based chain will be equal to $\pi$. We will need additional assumption
\begin{description}
\item[{\bf (H3)}] There exist such constants $\rho > 0, K_3 > 0$ that 
\begin{eqnarray*}
d_V(\delta_xQ_{\boldsymbol{\gamma}}^{1,n},\pi) \leq K_3 V(x) \rho^n 
\end{eqnarray*}
\end{description}

>>>>

In this section we perform the convergence analysis of the ULA algorithm. For the sake of simplicity we restrict our attention to the constant time step.
We shall use the notations $\mathbb N=\{1,2,\ldots\}$ and $\mathbb N_0=\mathbb N\cup\{0\}$.
By $H_k$, $k\in\mathbb N_0$,
we denote the normalized Hermite polynomial on $\mathbb R$, that is,
H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
\quad x\in\mathbb R.
For a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
$\mathbf{H}_\mathbf{k}$ denotes the normalized Hermite polynomial on $\mathbb R^d$, that is,
\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
In what follows, we also use the notation
$|\mathbf{k}|=\sum_{i=1}^d k_i$ for $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
and we set $\mathcal G_p=\sigma(Z_1,\ldots,Z_p)$, $p\in\mathbb N$, and $\mathcal G_0=\mathrm{triv}$.

@@

M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{p=N+1}^{N+n}\omega_{p,n}^{N}\sum_{l=N+1}^{p}a_{p,l,\mathbf{k}}(X_{l-1})\mathbf{H}_\mathbf{k}(Z_{l})
\label{eq:29032018a5}\\
&=\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{l=N+1}^{N+n}\bar{a}_{l,\mathbf{k}}(X_{l-1})\mathbf{H}_\mathbf{k}(Z_{l})

@@

with \(\|\mathbf{k}\|=\max_{i} k_i\) and
\bar{a}_{l,\mathbf{k}}(x)
& =\sum_{p=l}^{N+n}\omega_{p,n}^{N}a_{p,l,\mathbf{k}}(x) =\mathsf{E}\left[\left.\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right)\mathbf{H}_\mathbf{k}\left(Z_{l}\right)\right|X_{l-1}=x\right].
For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we shall be interested in the Mean Squared Error (MSE),
which can be decomposed as the sum of the squared  bias
and the  variance:
\begin{eqnarray}\label{eq:29032018a4}
\mathrm{MSE}\left[\rho(f)\right]
&=&\mathsf{E}\left[(\rho(f)-\pi(f))^2\right]
&=&\left(\mathsf{E}[\rho(f)]-\pi(f)\right)^2
+\mathsf{Var}[\rho(f)].

@@

Our analysis will be done under the following two assumptions.

\begin{description}
\item[(H1)][\textbf{Lipschitz continuity}] Assume that $\nabla U$ is Lipschitz, that is, there exists $L_U\geq 0$ such that 
\begin{eqnarray*}
|\nabla U(x)-\nabla U(y)|\leq L_U|x-y|, \quad x,y\in \mathbb{R}^d.
\end{eqnarray*}
\item[(H2)][\textbf{Convexity outside a ball}] There exist $K_U>0,$ $M_U>0$ and $m_U>0$ such that for any $x\not\in B_{K_U}(0)	$ and any $y\in \mathbb{R}^d$ it holds 
\begin{eqnarray*}
\langle D^2U(x)y,y\rangle \geq m_U\|y\|^2, \quad \|D^3 U(y)\| \leq M_U. 
\end{eqnarray*}
\end{description}
First we analyse the biases of the estimators
$\pi_n^N(f)$ and $\pi_{K,n}^N(f).$
\subsection{Squared bias}
Due to the martingale transform structure
of $M_{K,n}^N(f)$,
we have
\[
\mathsf E\left[M_{K,n}^N(f)\right]=0,
\]
Hence both estimators
$\pi_n^N(f)$ and $\pi_{K,n}^N(f)$ have the same conditional bias.
Notice that this remains true also when we substitute
the coefficients $a_{p,l,\mathbf{k}}$ in~\eqref{eq:29032018a5}
with some independent approximations $\widehat a_{p,l,\mathbf{k}}.$ For a bounded Borel function~$f$,
we can estimate the conditional bias
similarly to the beginning of ~\cite[Section~4]{durmus:moulines:2017}:
\begin{align}
\left(\mathsf{E}[\pi_{K,n}^N(f)]-\pi(f)\right)^2
&=\left(\mathsf{E}[\pi_{n}^N(f)]-\pi(f)\right)^2
\label{eq:06042018a2}\\
&\le
\mathrm{osc}(f)^2
\sum_{p=N+1}^{N+n}
\omega_{p,n}^N\,
\|\delta_x Q_\gamma^{p}-\pi(\cdot)\|_{TV}^2,
\notag
\end{align}
where
$\mathrm{osc}(f):=\sup_{x\in\mathbb R^d}f(x)-\inf_{x\in\mathbb R^d}f(x)$,
$\|\mu-\nu\|_{\TV}$ denotes the total variation distance
between probability measures $\mu$ and $\nu$, that is,
$$
\|\mu-\nu\|_{TV}=\sup_{A\in\mathcal B(\mathbb R^d)}
|\mu(A)-\nu(A)|,
$$
$\pi(\cdot)$ denotes the probability measure on $\mathbb R^d$
with density $\pi$ of~\eqref{eq:stationary_distr};
for $\gamma>0$, the Markov kernel $R_\gamma$
is defined by
$$
\mathcal{R}_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\}dy,
$$
while, for $k,l\in\mathbb N$, $k\le l$,
the kernel $\mathcal{Q}_{\boldsymbol{\gamma}}^{k,l}$ is
$
\mathcal{Q}_{\boldsymbol{\gamma}}^{k,l}=\mathcal{R}_{\gamma_l}\cdots \mathcal{R}_{\gamma_k},
$
which, finally, provides the  measure
$\mathcal{Q}_{\boldsymbol{\gamma}}^{p}=\mathcal{Q}_{\boldsymbol{\gamma}}^{p,1}$
used in the right-hand side of~\eqref{eq:06042018a2}.
Different specific upper bounds for the squared bias
can be deduced from~\eqref{eq:06042018a2}
using results of Section~3 in~\cite{durmus:moulines:2017}
on bounds in the total variation distance. For the case of a constant time step $\gamma_1=\ldots=\gamma_p=\gamma$ it is known that under assumption {\bf (H1)} the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$. Moreover, we shall show in Section that  under {\bf (H1)} and {\bf (H2)}
\begin{eqnarray}
\label{eq:bias}
\|\delta_x\mathcal{Q}_{\boldsymbol{\gamma}}^{n}-\pi_\gamma\|_{TV} \leq C\rho^{\gamma n}\left(V(x) + \pi_\gamma(V)\right)
\end{eqnarray}
for  $V(x) = 1 + \|x\|^2,$ some $\rho\in (0,1)$ independent of $\gamma$ and an absolute constant $C>0.$
\subsection{Conditional variance}
An upper bound for the variance of the classical estimator~\eqref{eq:29032018a2}
under {\bf (H1)} and {\bf (H2)} is derived in Section.  In particular, if $\sum_{j=1}^\infty \gamma_j=\infty$ and $\sum_{j=1}^\infty \gamma^2_j<\infty,$ then for any bounded function $f,$
\begin{eqnarray}
\label{eq:var-mc}
\mathsf{Var}[\pi_n^N(f)]\lesssim 1/\Gamma_{N+2,N+n+1},
\end{eqnarray}
where $\lesssim$ stands for inequality up to a constant depending on constants $L_U,$ $m_U,$  $M_U$ and $\|f\|_\infty.$ For a constant time step,  we derive from \eqref{eq:var-mc} that
$\mathsf{Var}[\pi_n^N(f)]\lesssim 1/(n\gamma).$

As for the estimator~\eqref{eq:29032018a3}, 
it follows from~\eqref{eq:29032018a5}, Theorem~\ref{prop:29032018a1} and independence condition for the sequence $(Z_p)_{p \geq 1}$.
\begin{prop}
\label{cond_var_dec}
	For variance-reduced estimator~\eqref{eq:29032018a3}, defined for the Markov Chain~\eqref{eq:chain_gen} the following representation holds
\begin{equation}\label{eq:29032018a6}
\mathsf{Var}\left[\pi_{K,n}^{N}(f)\right]
%=\sum_{\min_i (k_i)\ge K+1}
=\sum_{\mathbf k\colon \|\mathbf k\|\ge K+1}
\sum_{l=N+1}^{N+n}
\mathsf{E}\left[\left|\bar a_{l,\mathbf{k}}(X_{l-1})\right|^{2}\right].
\end{equation}
\end{prop}
One of the main results of this paper is the following  upper bound for the right-hand sides of~\eqref{eq:29032018a6} in the case of ULA algorithm.
\begin{thm}\label{th:mr}
Assume {\bf (H1)} and {\bf (H2)}.
Suppose that $f$  is two times continuously differentiable
and it holds
\begin{eqnarray*}
|\partial^{\mathbf{k}} f(x)|\le  B_f, \quad
x\in\mathbb R^d
\end{eqnarray*}
for all \(\mathbf{k}\) with \(0<\|\mathbf{k}\|\leq  2.\) Let $(\gamma_k)_{k\in\mathbb N}$ be a sequence of positive
and non-increasing step sizes with $\sum_{k=1}^\infty \gamma_k=\infty$ and $\sum_{k=1}^\infty \gamma^2_k<\infty.$ 
Then it holds under 
\begin{eqnarray}
\label{eq:var-bound}
\mathsf{Var}\left[\pi_{K,n}^{N}(f)\right]\lesssim \frac{1}{\Gamma^2_{N+2,N+n+1}},
\end{eqnarray}
where \(\lesssim\) stands for inequality up to a constant not depending  on \(n\) and \(N.\)
\end{thm}
\begin{cor}

@@

Let us formulate and prove prove analogue of the previous theorem for inhomogeneous Markov chain with step size $\gamma_1,\ldots,\gamma_n,\ldots$. Note that under assumptions $\sum_{p=1}^{\infty} \gamma_p=\infty, \quad \sum_{p=1}^{\infty} \gamma^2_p<\infty$ and {\bf (H1)} the stationary distribution of ULA-based chain will be equal to $\pi$. We will need additional assumption
\begin{description}
\item[{\bf (H3)}] There exist such constants $\rho > 0, K_3 > 0$ that 
\begin{eqnarray*}
d_V(\delta_xQ_{\boldsymbol{\gamma}}^{1,n},\pi) \leq K_3 V(x) \rho^{\Gamma_{1,n}} 
\end{eqnarray*}
\end{description}
